Evaluating contact tracing and isolation to control pandemic potential influenza outbreaks

### 2025-04-15 inst/paper/main.tex
add refs for reassortment and mutations to paper
polish interventions intro paragraph and re-render pdf

There are a variety of interventions that can be enacted to try and prevent transmission of an infectious disease, including both pharmaceutical and non-pharmaceutical interventions (NPIs). Early in an outbreak NPIs are often used, as especially for outbreaks of a novel pathogen before pharmaceutical inventions are available. These interventions range in their specificity, from targeted contact tracing which aims to tackle specific transmission pathways, to closure of school and gathering, to regional or national lockdowns affecting everyone (ref). \\

Contact tracing, the process of finding the contacts of infected individuals is an often used NPI to try and curtail the growth of an epidemic. It can involve forward contact tracing, whereby infected individuals are asked who they have in contact with so their contacts can be followed-up and isolated or quarantined is needed; or backward contact tracing whereby infected individuals are asked who they have been in contact with prior to being infected to identify the source of infection and potentially forward trace from the source. Contact tracing and isolation is a targeted approach, like ring vaccination, that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts \citep{kucharskiEffectivenessRingVaccination2016} (ref). Isolation of traced individuals has proved widely effective in controlling different outbreaks with a variety of pathogens differing in their aetiological and epidemiological characteristics (SARS ref; other refs). However, it's important to remember that it's not a panacea for all outbreak scenarios. In epidemics or pandemics where the proportion of cases or contacts missed is high, either because surveillance is patchy or because the number of contacts exceeds the capacity of the system, the efficacy of contact tracing to contain an outbreak diminishes, or at worst, fails completely \citep{dhillonWhenContactTracing2018} (ref on limited capacity of tracing). \\

@@

The reproduction number ($R$) use to simulate transmission for each influenza subtype was 1.1, 1.5, 2.5 and 3.5. We chose these values and they represent either similar values to previously estimated $R$ values for influenza outbreaks \citep{fergusonStrategiesMitigatingInfluenza2006}, including H1N1 \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}, or they represent an inflated values of current estimates of H5N1 and H7N9 subtypes to evaluate the effectiveness of contact tracing and isolation in the event that pathogen evolution (e.g. recombination of subtypes (ref)) enhances the human-to-human transmissibility of these pathogens. Current estimates for H5N1 and H7N9 subtypes are below unity (i.e. $R < 1$) (\citealt{tannerPandemicPotentialAvian2015}; \citealt{Ward2024.12.11.24318702}, but see \citealt{yangDetectingHumanhumanTransmission2007a} for a H5N1 $R$ estimate above 1) so transmission is expected to stop without intervention (except in a extremely rare case of superspreading event). Therefore evaluating contact tracing and isolation requires simulating outbreaks that are likely to exponentially spread without control measures.

>>>>

A range of interventions can be deployed to reduce or prevent the transmission of infectious diseases, broadly classified into pharmaceutical and non-pharmaceutical interventions (NPIs). Early in an outbreak -- especially in the case of a novel pathogen -- NPIs are often the first line of defence, as pharmaceutical options may not yet be available. These interventions vary in scope and specificity, from targeted measures such as contact tracing and isolation, to broader actions like school closures, bans on gatherings, and regional or national lockdowns (ref). The success of such interventions in containing an outbreak before it becomes sustained and widespread depends on several factors, including the type of intervention, the delay before implementation, and the intervention’s effectiveness (Longini et al., 2005). Even for highly transmissible pathogens (e.g. $R>3$) with short generation times (e.g. $<3$ days), swift and effective interventions can achieve containment. \\

Contact tracing, the process of finding the contacts of infected individuals is an often used NPI to try and curtail the growth of an epidemic. It can involve forward contact tracing, whereby infected individuals are asked who they have in contact with so their contacts can be followed-up and isolated or quarantined is needed; or backward contact tracing whereby infected individuals are asked who they have been in contact with prior to being infected to identify the source of infection and potentially forward trace from the source. Contact tracing and isolation is a targeted approach, like ring vaccination, that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts \citep{kucharskiEffectivenessRingVaccination2016} (Whittaker et al., 2024) (ref). Isolation of traced individuals has proved widely effective in controlling different outbreaks with a variety of pathogens differing in their aetiological and epidemiological characteristics (SARS ref; other refs). However, it's important to remember that it's not a panacea for all outbreak scenarios. In epidemics or pandemics where the proportion of cases or contacts missed is high, either because surveillance is patchy or because the number of contacts exceeds the capacity of the system, the efficacy of contact tracing to contain an outbreak diminishes, or at worst, fails completely \citep{dhillonWhenContactTracing2018} (ref on limited capacity of tracing). \\

@@

The reproduction number ($R$) use to simulate transmission for each influenza subtype was 1.1, 1.5, 2.5 and 3.5. We chose these values and they represent either similar values to previously estimated $R$ values for influenza outbreaks \citep{fergusonStrategiesMitigatingInfluenza2006}, including H1N1 \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}, or they represent an inflated values of current estimates of H5N1 and H7N9 subtypes to evaluate the effectiveness of contact tracing and isolation in the event that pathogen evolution (e.g. genetic reassortment (Peacock et al., 2025) or mutations in receptor binding preference (Lin et al., 2024)) enhances the human-to-human transmissibility of these pathogens. Current estimates for H5N1 and H7N9 subtypes are below unity (i.e. $R < 1$) (\citealt{tannerPandemicPotentialAvian2015}; \citealt{Ward2024.12.11.24318702}, but see \citealt{yangDetectingHumanhumanTransmission2007a} for a H5N1 $R$ estimate above 1) so transmission is expected to stop without intervention (except in a extremely rare case of superspreading event). Therefore evaluating contact tracing and isolation requires simulating outbreaks that are likely to exponentially spread without control measures.
